---
title: "Exploration"
author: "Ibrahim Al-Hindi"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Libraries
library(tidyverse)
library(here)
library(lubridate)
library(ggmap)
library(sf)
library(leaflet)
library(geosphere)
library(geodist)
```

```{r}
# GTFSr Data
updates_raw <- read_csv(here("data/GTFSr/large/gtfsr_prod.trip_updates.SAMPLE.csv.csv"))
stops_raw <- read_csv(here("data/GTFS_feed_version_1157/large/gtfs_history_prod.stops.csv"))
routes_raw <- read_csv(here("data/GTFS_feed_version_1157/gtfs_history_prod.routes.csv"))
sites_raw <- read_csv(here("data/addinsight/addinsight_prod.btsites.csv.csv"))
site_stats_raw <- read_csv(here("data/addinsight/large/addinsight_prod.btsitestats.SAMPLE.csv.csv"))
```

## Data Wrangling

```{r}
# Updates
updates <- updates_raw %>%
  rename(trip_id = id,
         vehicle_id = label) %>%
  # The same trip on the same date for the same stop can have multiple delay updates,
  # therefore filter updates to the most recent timestamp
  group_by(route_id, direction_id, start_date, trip_id, stop_sequence, stop_id) %>%
  filter(timestamp == max(timestamp)) %>%
  ungroup() %>% 
  mutate(direction_id = factor(direction_id, levels = c(1, 0)),
         start_date = ymd(start_date),
         vehicle_id = as.character(vehicle_id),
         timestamp = as_datetime(timestamp),
         trip_id = as.character(trip_id),
         stop_id = as.character(stop_id),
         arrival_time = as_datetime(arrival_time),
         date = date(timestamp),
         wday = wday(timestamp, label = TRUE, abbr = TRUE, week_start = 1),
         hour = hour(timestamp),
         time_stamp_floor = floor_date(timestamp, unit = "hour")) %>%
  mutate(delay_mins = delay/60, .after = delay) %>%
  mutate(stop_direction = paste0(stop_id, "-", direction_id), .after = stop_id) %>% 
  mutate(route_direction = paste0(route_id, "-", direction_id), .after = route_id) %>%
  mutate(route_dir_hr = paste0(route_id, "-", direction_id, "-", hour), .after = route_direction) %>% 
  # Get stop to stop delay per group
  group_by(route_id, direction_id, start_date, trip_id) %>% 
  arrange(stop_sequence, .by_group = TRUE) %>%
  mutate(stop_pair_dir = paste0(lag(stop_id), "-", stop_id, "-", unique(direction_id)), .after = delay_mins) %>%
  mutate(stop_to_stop_delay = delay - lag(delay, default = 0), .after = stop_pair_dir) %>%
  ungroup() %>%
  mutate(stop_to_stop_delay_mins = stop_to_stop_delay/60, .after = stop_to_stop_delay)
  
rm(updates_raw)

# Routes
routes <- routes_raw %>%
  distinct(route_id, .keep_all = TRUE) %>% 
  select(route_id, route_long_name:route_type, route_color, routegroup) %>%
  filter(
    # filter to buses only
    route_type == 3,
    # no train substitutes or hail n' ride services
    !str_detect(tolower(route_desc), pattern = "train substitute service|hail")) %>%
  mutate(route_color = paste0("#", route_color))

rm(routes_raw)

# Stops
stops <- stops_raw %>%
  distinct(stop_id, .keep_all = TRUE) %>% 
  select(stop_id, stop_name:stop_lon) %>% 
  mutate(stop_id = as.character(stop_id))
  
rm(stops_raw)

# Sites
sites <- sites_raw %>% 
  distinct(id, .keep_all = TRUE)

rm(sites_raw)

# Sites Stats
site_stats <- site_stats_raw %>% 
  mutate(logtime = as_datetime(str_remove(logtime, " \\+10:30"), tz = "UTC")) %>% 
  select(-c(dms_update_ts, schema_name:ingest_day)) %>% 
  filter(between(as_date(logtime), as_date("2022-01-01"), as_date("2022-01-31"))) %>% 
  mutate(hour = hour(logtime)) %>% 
  arrange(logtime)

rm(site_stats_raw)

# Join updates with routes and stops
combined <- updates %>%
  inner_join(routes, by = "route_id") %>% 
  inner_join(stops, by = "stop_id")
```

## Exploration

### Updates

```{r}
## Multiple trip_id can occur using the same combination of variables in group_by,
## need start_time
combined %>% 
  group_by(route_id, start_date, vehicle_id) %>% 
  summarise(trips = n_distinct(trip_id)) %>% 
  arrange(-trips)
```

```{r}
# delay per date and weekday
combined %>% 
  group_by(date, wday) %>% 
  summarise(avg_delay = mean(delay_mins)) %>% 
  ggplot(aes(x = date, y = avg_delay)) +
  geom_line(size = 2, alpha = 0.2) +
  geom_line(aes(color = wday)) +
  scale_color_brewer(type = "qual")

# delay per weekday
combined %>% 
  group_by(wday) %>% 
  summarise(avg_delay = mean(delay_mins)) %>% 
  ggplot(aes(x = wday, y = avg_delay)) +
  geom_col()
```

Barely any delays

## Delays

### Delays per Stop per Direction

```{r}
stops_delays <- combined %>%
  # Remove exceptional cases
  filter(stop_to_stop_delay_mins < 60) %>% 
  group_by(stop_id, direction_id) %>% 
  summarise(avg_delay = mean(stop_to_stop_delay_mins),
            occurrences = n()) %>%
  ungroup() %>%
  # Only include stop with number of occurrences greater than the median, indicates
  # abundance of data and importance of stop
  filter(occurrences > median(occurrences)) %>%
  arrange(-avg_delay)

# Select top 5 delays per stop_id
top_stops_delays <- stops_delays %>%
  slice_head(n=5)

# Filter data to stops identified
stops_combined <- combined %>% 
  semi_join(top_stops_delays, by = c("stop_id", "direction_id"))

stops_combined %>% 
  ggplot(aes(x = timestamp, y = stop_to_stop_delay_mins, color = stop_direction)) +
  geom_point() +
  geom_line()
```

## Delays per Route per Direction

```{r}
routes_delays <- combined %>%
  filter(delay_mins < 40) %>% 
  group_by(route_id, direction_id) %>%
  summarise(avg_delay = mean(delay_mins),
            # A start_date can have several trip_id's, and a trip_id can happen in
            # several start_dates, so count unique combinations of each
            no_of_trips = n_distinct(start_date, trip_id),
            no_of_days = n_distinct(start_date),
            no_of_unique_trips = n_distinct(trip_id),
            route_name = unique(route_long_name),
            route_desc = unique(route_desc),
            route_group = unique(routegroup),
            route_color = unique(route_color)) %>%
  ungroup() %>% 
  filter(no_of_trips > median(no_of_trips)) %>%
  arrange(-avg_delay)

top_routes_delays <- routes_delays %>% 
  slice_head(n=5)

routes_combined <- combined %>% 
  semi_join(top_routes_delays, by = c("route_id", "direction_id"))

ggplot(routes_combined, aes(x = timestamp, y = stop_to_stop_delay_mins, color = route_direction)) +
  geom_point() +
  geom_line()
```

## Delays per Hour

```{r}
hourly_delays <- combined %>%
  group_by(hour) %>% 
  summarise(avg_delay = mean(delay_mins)) %>%
  ungroup() %>%
  arrange(hour)

ggplot(hourly_delays, aes(hour, avg_delay)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = seq(0,23)) +
  theme(panel.grid.minor = element_blank())
```

## Delays per Route per Direction per Hour

```{r}
routes_hourly_delays <- combined %>% 
  filter(delay_mins < 40) %>%
  group_by(route_id, direction_id, hour) %>% 
  summarise(route_direction = unique(route_direction),
            avg_delay = mean(delay_mins),
            no_of_trips = n_distinct(start_date, trip_id),
            no_of_days = n_distinct(start_date),
            no_of_unique_trips = n_distinct(trip_id),
            route_name = unique(route_long_name),
            route_desc = unique(route_desc),
            route_group = unique(routegroup),
            route_color = unique(route_color)) %>% 
  ungroup() %>%
  arrange(-avg_delay)

top_routes_hourly_delays <- routes_hourly_delays %>%
  filter(no_of_trips > median(no_of_trips)) %>% 
  slice_head(n=5)

routes_hourly_combined <- combined %>% 
  semi_join(top_routes_hourly_delays, by = c("route_id", "direction_id", "hour"))

ggplot(routes_hourly_combined, aes(x = timestamp, y = stop_to_stop_delay_mins, color = route_dir_hr)) +
  geom_point() +
  geom_line()

# Per each hour
top_delays_per_hour <- routes_hourly_delays %>% 
  group_by(hour) %>% 
  slice_max(order_by = avg_delay, n = 1) %>% 
  ungroup()

hourly_delays_combined <- combined %>% 
  semi_join(top_delays_per_hour, by = c("route_id", "direction_id", "hour"))

ggplot(top_delays_per_hour, aes(x = hour, y = avg_delay)) +
  geom_line() +
  geom_text(aes(label = route_direction), color = "red")

```

Continue with the top 5 routes-directions identified earlier.


## Identify hour and trip_id to use per top 5 routes

```{r}
# Identify hour with largest delays per route
trip_hr_per_route <- routes_combined %>% 
  group_by(route_id, direction_id, hour) %>% 
  summarise(avg_delay = mean(delay_mins),
            no_of_trips = n_distinct(start_date, trip_id),
            no_of_days = n_distinct(start_date),
            no_of_unique_trips = n_distinct(trip_id),
            route_name = unique(route_long_name),
            route_desc = unique(route_desc),
            route_group = unique(routegroup),
            route_color = unique(route_color)) %>%
  ungroup() %>% 
  group_by(route_id, direction_id) %>%
  filter(no_of_trips > mean(no_of_trips)) %>%
  filter(avg_delay == max(avg_delay)) %>% 
  ungroup()

# Extract all trips per route, direction and hour identified
route_trips <- routes_combined %>% 
  semi_join(trip_hr_per_route, by = c("route_id", "direction_id", "hour")) %>% 
  group_by(route_id, direction_id, hour) %>% 
  summarise(trip_id = unique(trip_id)) %>% 
  ungroup()

# For each trip_id identified, get the specific trip with maximum stops
set.seed(1)
trips_max_stops <- routes_combined %>% 
  semi_join(route_trips, by = c("route_id", "direction_id", "trip_id")) %>%
  group_by(route_id, start_date, trip_id) %>% 
  summarise(no_of_stops = n_distinct(stop_sequence)) %>%
  ungroup() %>% 
  group_by(route_id) %>% 
  slice_max(order_by = no_of_stops, n = 1) %>%
  # Route 403 has several trips with the same number of max stops, get one randomly
  slice_sample(n = 1) %>% 
  ungroup()

# Get full details of identified trips
trips_to_map <- routes_combined %>% 
  semi_join(trips_max_stops, by = c("route_id", "start_date", "trip_id")) %>% 
  # routes 224M and 403 have the same color, change one of them
  mutate(route_color = ifelse(route_id == "224M", "#FFC733", route_color))
```

# Map

A function will be built to extract sites and construct map based on routes provided

```{r}
extract_sites <- function(...) {
  
  trips <- c(...)
  
  if (length(trips) == 1 && trips == "all") {
    trips <- unique(trips_to_map$route_id)
  } 
  
  if (any(!trips %in% unique(trips_to_map$route_id))) {
    stop("Route(s) is not an examined route. Type 'all' to select all routes")
  }
  
  # filter routes to pass into map
  selected_routes <- trips_to_map %>% 
  filter(route_id %in% trips)
  
  # Get sites close to stops
  stops_coords <- selected_routes %>% 
    select(stop_id, stop_lat, stop_lon) %>% 
    distinct() %>% 
    rename(lat = stop_lat, lon = stop_lon)

  sites_coords <- sites %>% 
    select(id, longitude, latitude) %>% 
    rename(lat = latitude, lon = longitude)
  
  # calculate all pairwise distances between stops and sites
  stops_sites_dist <- geodist(x = sites_coords, y = stops_coords, measure = "haversine")
  
  colnames(stops_sites_dist) <- stops_coords$stop_id
  
  stops_sites_dist <- stops_sites_dist %>% 
    as_tibble() %>% 
    mutate(site = sites$id, .before = everything())
  
  # Keep sites within 300m of any stop and assign as global variable
  sites_300m <- stops_sites_dist %>% 
    filter(if_any(c(everything(), -site), ~ .x < 300))
  
  sites_map <- sites %>% 
    semi_join(sites_300m, by = c("id" = "site"))
  
  # Set color palette for leaflet legend
  pal <- colorFactor(palette = unique(selected_routes$route_color),
                              domain = selected_routes$route_id)
  
  selected_routes %>% 
    leaflet() %>% 
    addTiles() %>% 
    addCircles(lng = ~ stop_lon,
               lat = ~ stop_lat,
               color = ~ route_color,
               popup = ~ stop_id,
               radius = 50,
               fillOpacity = 1) %>% 
    addCircles(data = sites_map,
               lng = ~ longitude,
               lat = ~ latitude,
               popup = ~ as.character(id),
               color = "green",
               fillOpacity = 1) %>% 
    addLegend(position = "bottomleft", pal = pal, values = ~ route_id, title = "Routes")
}

extract_sites("all")
```

A data frame containing each route and its associated steps will be created

```{r}
# Initialize list to contain data frames for each route then combine at the end
routes_sites <- vector("list", n_distinct(trips_to_map$route_id))

for (i in seq_along(unique(trips_to_map$route_id))) {
  
  selected_route <- trips_to_map %>% 
  filter(route_id == unique(trips_to_map$route_id)[i])
  
  stops_coords <- selected_route %>% 
    select(stop_id, stop_lat, stop_lon) %>% 
    distinct() %>% 
    rename(lat = stop_lat, lon = stop_lon)
  
  sites_coords <- sites %>% 
    select(id, longitude, latitude) %>% 
    rename(lat = latitude, lon = longitude)
  
  stops_sites_dist <- geodist(x = sites_coords, y = stops_coords, measure = "haversine")
  
  colnames(stops_sites_dist) <- stops_coords$stop_id
  
  stops_sites_dist <- stops_sites_dist %>% 
    as_tibble() %>% 
    mutate(site = sites$id, .before = everything())
  
  sites_300m <- stops_sites_dist %>% 
    filter(if_any(c(everything(), -site), ~ .x < 300))
  
  routes_sites[[i]] <- sites %>% 
    semi_join(sites_300m, by = c("id" = "site")) %>% 
    mutate(route_id = unique(selected_route$route_id))
}

routes_sites <- bind_rows(routes_sites)
```

# Sites

```{r}
ggplot(site_stats, aes(probecount, avgduration))
```

```{r}
# Calculate average probecount and duration per site
site_count <- site_stats %>% 
  group_by(siteid) %>% 
  summarise(avg_count = mean(probecount),
            avg_duration = mean(avgduration)) %>% 
  ungroup() %>% 
  arrange(-avg_count)

ggplot(site_count, aes(avg_count)) +
  geom_histogram()

ggplot(site_count, aes(x = avg_count, y = avg_duration)) +
  geom_point()

# Calculate average probecount and duration per hour
site_hour_count <- site_stats %>% 
  group_by(hour) %>% 
  summarise(avg_count = mean(probecount),
            avg_duration = mean(avgduration)) %>% 
  ungroup() %>% 
  arrange(-avg_count)

ggplot(site_hour_count, aes(x = hour, y = avg_count)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = seq(0,23)) +
  theme(panel.grid.minor = element_blank())

hourly_delays_congestion <- site_hour_count %>% 
  left_join(hourly_delays, by = "hour") %>% 
  pivot_longer(cols = -hour, names_to = "measure")

ggplot(hourly_delays_congestion, aes(x = hour, y = value)) +
  geom_line() +
  scale_x_continuous(breaks = seq(0,23)) +
  theme(panel.grid.minor = element_blank()) +
  facet_grid(vars(measure), scales = "free_y")

# Calculate average probecount and duration per site per hour
site_hour_combo_count <- site_stats %>%
  group_by(siteid, hour) %>% 
  summarise(avg_count = mean(probecount),
            avg_duration = mean(avgduration)) %>% 
  ungroup() %>% 
  arrange(-avg_count)

ggplot(site_hour_combo_count, aes(x = avg_count)) +
  geom_histogram()
```
























































