```{r libraries}
# Libraries
library(tidyverse)
library(here)
library(lubridate)
library(leaflet)
library(hms)
library(janitor)
library(broom)
```

```{r data}
# Data

## Holidays dates
holidays_raw <- read_csv(here("data/addinsight/addinsight_prod.btspecialdays.csv.csv"))

## Links information
links_raw <- read_csv(here("data/addinsight/addinsight_prod.btlinks.csv.csv"))

## List of links on segment and their geometries
links_segment_raw <- read_csv(here("data/Mariod-Rd/Marion Rd Links.csv"))

## Stops information
full_stops_raw <- read_csv(here("data/GTFS_feed_version_1157/large/gtfs_history_prod.stops.csv"))

## Routes, trips, and stops on segment
segment_routes_stops_raw <- read_csv(here("data/Mariod-Rd/Marion Rd Routes Stops.csv"))

## GTFS-R Trip Updates
updates_raw <- read_csv(here("data/Mariod-Rd/trip_updates_MarionRd_Mar.csv"))

## Link 5 minute statistics
link_stats_raw <- read_csv(here("data/Mariod-Rd/link_5minstats_Mar.csv"))
```

# Data Wrangling and Exploration

## Holidays

```{r}
holidays <- unique(holidays_raw$date)
rm(holidays_raw)
```

## Links Segment

```{r}
# Limit updates to end of period examined and get most recent update per link
links <- links_raw %>%
  filter(date(dms_update_ts) < "2022-04-01") %>%
  group_by(id) %>%
  filter(dms_update_ts == max(dms_update_ts)) %>%
  ungroup() %>%
  select(id, originid:direction) %>%
  distinct()

links_segment <- links_segment_raw %>%
  rename_with(tolower) %>%
  mutate(
    latitude = round(latitude, 5),
    longitude = round(longitude, 5),
    # Extract name of locations where link starts and ends
    start_loc = str_extract(name, "(?<=- ).*(?= to)"),
    end_loc = str_extract(name, "(?<=to ).*")) %>%
  group_by(linkid) %>%
  mutate(
    start_lat = first(latitude, order_by = ordernumber),
    end_lat = last(latitude, order_by = ordernumber)) %>%
  ungroup() %>%
  # Length in links_segment_raw is not correct, join with links to get correct lengths and direction
  select(-id, -length) %>% 
  left_join(links, by = c("linkid" = "id")) %>% 
  # Rename SWB direction to SB
  mutate(direction = recode(direction, SWB = "SB"))

# Get only first entry per link
links_segment_unq <- links_segment %>%
  arrange(linkid, ordernumber) %>%
  distinct(linkid, .keep_all = TRUE)

directions <- sort(unique(links_segment_unq$direction))

rm(links_raw)
rm(links_segment_raw)
```

## Full Stops and Route-Trips-Stops

```{r}
segment_routes_stops <- clean_names(segment_routes_stops_raw)
  
segment_routes <- segment_routes_stops %>% 
  pull(route_id) %>% 
  unique()

segment_stops <- segment_routes_stops %>% 
  pull(stop_id_gtfs_history_prod_stops_csv_2) %>% 
  unique()

# Join stops information to stops on the segment
stops_segment <- full_stops_raw %>%
  filter(stop_id %in% segment_stops) %>%
  select(stop_id, stop_name, stop_desc, stop_lat, stop_lon) %>%
  distinct(stop_id, .keep_all = TRUE) %>% 
  # Remove non-bus stops
  filter(str_detect(stop_name, "West|East")) %>% 
  # If stop is on East side, that means trip SB, and vice versa
  mutate(direction = if_else(str_detect(stop_name, "East"), "SB", "NB"))

rm(segment_routes_stops_raw)
rm(full_stops_raw)
```

## Updates

```{r}
# Keep only updates for stops on segment and remove weekends and public holidays
updates <- updates_raw %>%
  rename(
    trip_id = id,
    vehicle_id = label) %>%
  mutate(
    start_date = ymd(start_date),
    timestamp = as_datetime(timestamp, tz = "Australia/Adelaide"),
    arrival_time = as_datetime(arrival_time, tz = "Australia/Adelaide")) %>%
  inner_join(select(stops_segment, stop_id, direction), by = "stop_id") %>%
  filter(
    !start_date %in% holidays,
    !wday(start_date, label = TRUE) %in% c("Sat", "Sun"))

# Clean updates and get time to each stop for error detection
updates_stop_times <- updates %>% 
  group_by(start_date, trip_id) %>%
  arrange(stop_sequence, .by_group = TRUE) %>%
  # Remove trips where all the stops have a delay over 2400 (40 minutes), 
  # or all the stops are over 900 (15 minutes) early. Most likely error due to
  # entering the information later. This is done to prevent
  # incorrect analysis since they will be in the wrong time period
  filter(!(all(delay > 2400) |
           all(delay < -900))) %>%
  # Remove one-off large jumps in between two observations
  filter(!(delay - lag(delay, order_by = stop_sequence, default = 0) > 1000 &
             delay - lead(delay, order_by = stop_sequence, default = 0) > 1000)) %>%
  # Remove any observations that have an arrival_time later than any following
  # arrival_time in the trip AND the timestamp is earlier than any following timestamps in the trip.
  # This ensures the most recent timestamp is preferred when discrepancy occurs
  filter(!(as.numeric(arrival_time) > order_by(-stop_sequence, cummin(as.numeric(arrival_time))) &
             as.numeric(timestamp) == order_by(-stop_sequence, cummin(as.numeric(timestamp))))) %>%
  # Remove if the arrival_time is less than the previous arrival_times AND the timestamp
  # is older than the previous timestamps
  filter(!(arrival_time < order_by(stop_sequence, cummax(as.numeric(arrival_time))) &
             timestamp < order_by(stop_sequence, cummax(as.numeric(timestamp))))) %>%
  # If arrival_time of a stop is less than prior stops but they all have the same
  # timestamp, it's not possible to know which is correct. Assume earlier stop_sequence
  # is correct since it is closer when the update is made
  filter(!(timestamp == order_by(stop_sequence, cummax(as.numeric(timestamp))) &
             arrival_time < order_by(stop_sequence, cummax(as.numeric(arrival_time))))) %>%
  # If after the above, two consecutive stops have the same arrival_time, remove
  # the one with an older timestamp
  filter(!(arrival_time == lead(arrival_time, 
                                order_by = stop_sequence,
                                default = ymd("9999-01-01")) &
             timestamp < lead(timestamp, 
                              order_by = stop_sequence,
                              default = ymd("9999-01-01")))) %>%
  # If two stops have the same arrival_time and same time_stamp, remove the one
  # with a higher stop sequence. Make sure only 2 stops and not more
  filter(!(arrival_time != lead(arrival_time, 
                                order_by = stop_sequence,
                                default = ymd("9999-01-01")) &
             arrival_time == lag(arrival_time, 
                                 order_by = stop_sequence,
                                 default = ymd("0000-01-01")) &
             timestamp == lag(timestamp, 
                              order_by = stop_sequence,
                              default = ymd("0000-01-01")))) %>%
  # Remove trips with multiple repeating arrival_times
  filter(!n_distinct(arrival_time) < n()) %>%
  mutate(
    to_stop_time = as.numeric(arrival_time - lag(arrival_time), units = "secs"),
    first_stop = first(stop_id, order_by = stop_sequence),
    last_stop = last(stop_id, order_by = stop_sequence)) %>% 
  ungroup() %>%
  mutate(to_stop_time = replace_na(to_stop_time, 0))

paste0("Percentage of errors: ", round((nrow(updates) - nrow(updates_stop_times)) / nrow(updates) * 100, 2), "%")

# Take top occurring stops pair for each direction
top_first_last_stops <- updates_stop_times %>% 
  distinct(start_date, trip_id, direction, first_stop, last_stop) %>%
  group_by(direction) %>% 
  count(first_stop, last_stop, sort = TRUE) %>% 
  slice_max(order_by = n) %>% 
  ungroup()

# Get trips with only most occurring stops pair per direction identified
updates_stop_times <- updates_stop_times %>%
  semi_join(top_first_last_stops, by = c("direction", "first_stop", "last_stop"))

# Get total time per trip within segment
updates_trip_time <- updates_stop_times %>% 
  group_by(start_date, trip_id) %>%
  arrange(stop_sequence, .by_group = TRUE) %>%
  mutate(
    trip_time = as.numeric((last(arrival_time) -first(arrival_time)), units = "secs"),
    delay_diff = abs(last(delay) - first(delay))) %>% 
  ungroup() %>%
  # If the delay_diff is greater than 600 (10 minutes), most likely an error. Remove
  filter(delay_diff < 600) %>% 
  arrange(start_date, trip_id, stop_sequence)

# One row per trip corresponding to highest delay in the trip. For error detection using plots
trip_time_delay <- updates_trip_time %>% 
  group_by(start_date, trip_id) %>% 
  filter(abs(delay) == max(abs(delay))) %>%
  ungroup() %>% 
  distinct(start_date, trip_id, .keep_all = TRUE)

# One row per trip corresponding to highest to_stop_time in the trip. For error detection using plots
trip_time_stop <- updates_trip_time %>% 
  group_by(start_date, trip_id) %>% 
  filter(abs(to_stop_time) == max(abs(to_stop_time))) %>%
  ungroup() %>% 
  distinct(start_date, trip_id, .keep_all = TRUE)

# For further analysis, take first stop only from each trip because the arrival_time
# of the first stop will be considered as the trip start time used as basis for aggregation later.
# Also add whether each trip occurrs during either of the peak times
trip_times <- updates_trip_time %>% 
  group_by(start_date, trip_id) %>% 
  filter(stop_sequence == first(stop_sequence, order_by = stop_sequence)) %>%
  ungroup() %>% 
  distinct(start_date, trip_id, .keep_all = TRUE) %>% 
  mutate(
    rush = case_when(
      as_hms("06:00:00") <= as_hms(arrival_time) & as_hms(arrival_time) < as_hms("10:00:00") ~ "morning",
      as_hms("15:00:00") <= as_hms(arrival_time) & as_hms(arrival_time) < as_hms("19:00:00") ~ "evening",
      TRUE ~ "neither"),
    rush = factor(rush, levels = c("morning", "evening", "neither")))

trip_times_peaks <- trip_times %>% 
  filter(rush %in% c("morning", "evening"))

# Join with trip_times to only have trips analysed
stop_times <- updates_stop_times %>% 
  semi_join(trip_times, by = c("start_date", "trip_id"))

# Update stops_segment to only pairs of stops needed
stops_segment <- stops_segment %>% 
  filter(stop_id %in% c(top_first_last_stops$first_stop, top_first_last_stops$last_stop))

rm(updates_raw)
```

## Data Validation

```{r}
ggplot(updates_stop_times, aes(to_stop_time)) +
  geom_histogram() +
  theme_bw()

ggplot(updates_stop_times, aes(to_stop_time)) +
  geom_boxplot() +
  theme_bw()

ggplot(updates_stop_times, aes(delay)) +
  geom_histogram() +
  theme_bw()

ggplot(updates_stop_times, aes(delay)) +
  geom_boxplot() +
  theme_bw()

ggplot(updates_stop_times, aes(to_stop_time, delay)) +
  geom_point() +
  theme_bw()

ggplot(updates_trip_time, aes(trip_time)) +
  geom_histogram() +
  theme_bw()

ggplot(updates_trip_time, aes(trip_time)) +
  geom_boxplot() +
  theme_bw()

ggplot(trip_time_delay, aes(trip_time, delay)) +
  geom_point() +
  facet_wrap(vars(direction)) +
  theme_bw()

ggplot(trip_time_stop, aes(trip_time, to_stop_time)) +
  geom_point() +
  facet_wrap(vars(direction)) +
  theme_bw()

ggplot(trip_time_delay, aes(trip_time, delay_diff)) +
  geom_point() +
  facet_wrap(vars(direction)) +
  theme_bw()

ggplot(trip_time_delay, aes(delay_diff, color = direction)) +
  geom_boxplot() +
  theme_bw()

ggplot(trip_time_delay, aes(trip_time, delay_diff, color = direction)) +
  geom_point() +
  theme_bw()

ggplot(trip_time_stop, aes(trip_time, to_stop_time, color = direction)) +
  geom_point() +
  theme_bw()

ggplot(stop_times, aes(to_stop_time)) +
  geom_histogram() +
  theme_bw()

ggplot(stop_times, aes(to_stop_time)) +
  geom_boxplot() +
  theme_bw()

ggplot(stop_times, aes(delay)) +
  geom_histogram() +
  theme_bw()

ggplot(stop_times, aes(delay)) +
  geom_boxplot() +
  theme_bw()

ggplot(stop_times, aes(to_stop_time, delay)) +
  geom_point() +
  theme_bw()

ggplot(trip_times_peaks, aes(direction, trip_time)) +
  geom_boxplot() +
  facet_wrap(vars(rush)) +
  theme_bw()

ggplot(trip_times_peaks, aes(trip_time, color = direction)) +
  geom_density() +
  facet_wrap(vars(rush)) +
  theme_bw()
```

# Getting Information for Analysis

## Identify Routes

Routes were identified by examining the [network map](https://www.adelaidemetro.com.au/__data/assets/pdf_file/0009/824247/Adelaide-Metro-network-map.pdf).

All the routes were overlayed on a map on Tableau and routes on South Road were selected and exported.

## Identify Stops

Same as above

```{r}
# Plot on map to check the stops are all on South Rd
stops_segment %>%
    leaflet() %>% 
    addTiles() %>% 
    addCircles(lat = ~stop_lat, lng = ~stop_lon, label = ~ stop_id)
```

## Identify Links

### Direction 1

```{r}
# The going back and forth between links_dir process is for the purpose of
# producing links_disabled 

links_dir1 <- links_segment_unq %>%
  # Build map from south to north
  filter(direction == directions[1]) %>% 
  arrange(start_lat)

links_dir1_enabled <- links_dir1 %>%
  filter(enabled)

# Start and end at same locations as bus stops by referencing map above
dir1_start_loc <- "peter"
dir1_end_loc <- "glengyle"

# Retrieve sequence of links composing segment
links_dir1_enabled <- links_dir1_enabled[min(which(str_starts(tolower(links_dir1_enabled$start_loc),
                                                                  tolower(dir1_start_loc)))):
                                               max(which(str_starts(tolower(links_dir1_enabled$end_loc),
                                                                    tolower(dir1_end_loc)))),] %>% 
  group_by(start_loc) %>%
  filter(length == min(length)) %>%
  ungroup()

# Obtain shape for each link
links_dir1_map <- links_segment %>% 
  semi_join(links_dir1_enabled, by = "linkid") %>% 
  arrange(start_lat, ordernumber)

# Create map

links_colors <- c("green","red")

n <- 1

map_dir1 <- leaflet() %>% 
  addTiles()

# Iterate over links to add polyline of each link to the map
for (link in unique(links_dir1_map$linkid)) {
  map_df <- links_dir1_map %>% 
    filter(linkid == link)
  
  map_dir1 <- map_dir1 %>% 
    addPolylines(data = map_df, lng = ~longitude, lat = ~latitude, color = links_colors[n],
                 label = ~linkid, opacity = 0.3)
  
  # Switch colors
  if (n == 1) {
    n <- 2
  } else {
    n <- 1
  }
}

map_dir1
```

### Direction 2

```{r}
# Same process as direction 1

links_dir2 <- links_segment_unq %>%
  filter(direction == directions[2]) %>% 
  arrange(-start_lat)

links_dir2_enabled <- links_dir2 %>%
  filter(enabled)

dir2_start_loc <- "glengyle"
dir2_end_loc <- "peter"

links_dir2_enabled <- links_dir2_enabled[min(which(str_starts(tolower(links_dir2_enabled$start_loc),
                                                                  tolower(dir2_start_loc)))):
                                               max(which(str_starts(tolower(links_dir2_enabled$end_loc),
                                                                    tolower(dir2_end_loc)))),] %>% 
  group_by(start_loc) %>%
  filter(length == min(length)) %>%
  ungroup()

links_dir2_map <- links_segment %>% 
  semi_join(links_dir2_enabled, by = "linkid") %>% 
  arrange(-start_lat, ordernumber)

links_colors <- c("green","red")

n <- 1

map_dir2 <- leaflet() %>% 
  addTiles()

for (link in unique(links_dir2_map$linkid)) {
  map_df <- links_dir2_map %>% 
    filter(linkid == link)
  
  map_dir2 <- map_dir2 %>% 
    addPolylines(data = map_df, lng = ~longitude, lat = ~latitude, color = links_colors[n],
                 label = ~linkid, opacity = 0.3)

  if (n == 1) {
    n <- 2
  } else {
    n <- 1
  }
}

map_dir2
```

### Missing Links

No missing links

```{r}
# DF for missing links imputation
missing_links_impute <- tibble(from = character(),
                               to = character(),
                               linkid = numeric(),
                               prior_linkid = numeric(),
                               post_linkid = numeric(),
                               direction = character())
```

### Both directions together

```{r}
# Plot links from both directions
map_both <- leaflet() %>% 
  addTiles()

# Add links first direction
for (link in unique(links_dir1_map$linkid)) {
  map_df <- links_dir1_map %>% 
    filter(linkid == link)
  
  map_both <- map_both %>% 
    addPolylines(data = map_df, lng = ~longitude, lat = ~latitude, color = "green", 
                 opacity = 0.3, label = ~linkid)
}

# Add links second direction
for (link in unique(links_dir2_map$linkid)) {
  map_df <- links_dir2_map %>% 
    filter(linkid == link)
  
  map_both <- map_both %>% 
    addPolylines(data = map_df, lng = ~longitude, lat = ~latitude, color = "red",
                 opacity = 0.3, label = ~linkid)
}

map_both

# Save linkid's for both directions, the missing links, and extra links needed to impute missing links
links_all <- unique(c(links_dir1_enabled$linkid, links_dir2_enabled$linkid,
                missing_links_impute$linkid,missing_links_impute$prior_linkid, 
                missing_links_impute$post_linkid))
```

### Link Stats

Links stats are retrieved from AWS Athena database using the links identified to construct the segments and the links needed to impute missing links stats

```{r}
link_stats <- link_stats_raw %>% 
  select(logtime:avgdensity) %>%
  mutate(logtime = ymd_hms(str_remove(logtime, " \\+10:30"), tz = "Australia/Adelaide")) %>%
  filter(
    !as_date(logtime) %in% holidays,
    !wday(logtime, label = TRUE) %in% c("Sat", "Sun")) %>% 
  distinct()

# Join link_stats with links_segment to get direction and length
link_stats <- link_stats %>%
  left_join(select(links_segment_unq, linkid, direction, length)) %>%
  # Calculate speed km/h
  mutate(speed = round((length / 1000) / (tt / 3600))) %>% 
  # Remove speeds over 150 km/h as most likely errors
  filter(speed < 150)

summary(link_stats)

rm(link_stats_raw)
```

#### Data Validation

```{r}
ggplot(link_stats, aes(tt)) +
  geom_histogram() +
  theme_bw()

ggplot(link_stats, aes(tt)) +
  geom_boxplot() +
  theme_bw()

ggplot(link_stats, aes(speed)) +
  geom_histogram() +
  theme_bw()

ggplot(link_stats, aes(speed)) +
  geom_boxplot() +
  theme_bw()

ggplot(link_stats, aes(factor(length), tt)) +
  geom_boxplot() +
  theme_bw()

ggplot(link_stats, aes(speed, tt)) +
  geom_point() +
  theme_bw()

ggplot(link_stats, aes(factor(length), speed)) +
  geom_boxplot() +
  theme_bw()
```

### Check for links with no stats

```{r}
links_with_stats <- links_segment %>% 
  filter(linkid %in% c(links_dir1_enabled$linkid, links_dir2_enabled$linkid,
                       missing_links_impute$linkid)) %>% 
  semi_join(link_stats, by = "linkid")

links_no_stats <- links_segment %>% 
  filter(linkid %in% missing_links_impute$linkid) %>%  
  anti_join(link_stats, by = "linkid")
  
map_links_with_stats <- leaflet() %>% 
  addTiles()

# Draw links with stats in green
for (link in unique(links_with_stats$linkid)) {
  map_df <- links_with_stats %>% 
    filter(linkid == link)
  
  map_links_with_stats <- map_links_with_stats %>% 
    addPolylines(data = map_df, lng = ~longitude, lat = ~latitude, color = "green",
                 opacity = 1, label = ~linkid)
}

# Draw links with no stats in red
for (link in unique(links_no_stats$linkid)) {
  map_df <- links_no_stats %>% 
    filter(linkid == link)
  
  map_links_with_stats <- map_links_with_stats %>% 
    addPolylines(data = map_df, lng = ~longitude, lat = ~latitude, color = "red",
                 opacity = 1, label = ~linkid)
}

map_links_with_stats
```

### Generate stats for missing links

No need

```{r}
# Calculate total links travel time per logtime and direction
links_times <- link_stats %>%
  group_by(logtime, direction) %>%
  summarise(links_time = sum(tt)) %>%
  ungroup() %>%
  mutate(
    day = date(logtime),
    time = as_hms(logtime),
    hour = hour(logtime),
    rush = case_when(
      between(hour, 6, 9) ~ "morning",
      between(hour, 15, 18) ~ "evening",
      TRUE ~ "neither"),
    rush = factor(rush, levels = c("morning", "evening", "neither"))) %>%
  relocate(day:rush, .after = logtime) %>% 
  select(-logtime)
```

```{r}
ggplot(links_times, aes(links_time)) +
  geom_histogram() +
  theme_bw()

ggplot(links_times, aes(links_time)) +
  geom_boxplot() +
  theme_bw()
```

# Analysis

```{r}
# Split trips into 5 minute intervals average the trips times per interval
trip_times_agg <- trip_times %>% 
  mutate(arrival_time = ceiling_date(arrival_time, "5 minutes")) %>% 
  group_by(arrival_time, direction) %>% 
  summarise(
    number_buses = n(),
    bus_time = round(mean(trip_time))) %>%
  ungroup() %>% 
  mutate(
    day = date(arrival_time),
    time = as_hms(arrival_time),
    hour = hour(time),
    rush = case_when(
      between(hour, 6, 9) ~ "morning",
      between(hour, 15, 18) ~ "evening",
      TRUE ~ "neither"),
    rush = factor(rush, levels = c("morning", "evening", "neither"))) %>% 
  relocate(day:rush, .after = arrival_time) %>% 
  select(-arrival_time)
```

## Compare bus travel times to links travel times to validate each other

```{r}
# Combine links times and bus trip times
travel_times <- trip_times_agg %>%
  inner_join(links_times) %>% 
  mutate(travel_diff = bus_time - links_time)

travel_times_piv <- travel_times %>% 
  pivot_longer(cols = bus_time:links_time, names_to = "type", values_to = "travel_time")
```

```{r}
# Travel time distributions
travel_times_piv %>%
  ggplot(aes(travel_time, fill = type)) +
  geom_density(alpha = 0.5) +
  theme_bw() +
  theme(
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    axis.ticks.y = element_blank(),
    legend.title = element_blank())

travel_times_piv %>% 
  ggplot(aes(travel_time, fill = type)) +
  geom_density(alpha = 0.5) +
  facet_wrap(vars(direction)) +
  theme_bw() +
  theme(
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    axis.ticks.y = element_blank(),
    legend.title = element_blank())

# Travel difference distributions
travel_times %>% 
  ggplot(aes(travel_diff)) +
  geom_histogram() +
  theme_bw()

travel_times %>% 
  ggplot(aes(travel_diff)) +
  geom_histogram() +
  facet_wrap(vars(direction)) +
  theme_bw()

# Direction 1
travel_times_dir1 <- travel_times %>% 
  filter(direction == directions[1])

travel_times_piv_dir1 <- travel_times_piv %>% 
  filter(direction == directions[1])

travel_times_piv_dir1 %>%
  ggplot(aes(time, travel_time, color = type)) +
  geom_line() +
  facet_wrap(vars(day)) +
  ggtitle(directions[1]) +
  theme_bw() +
  theme(
    legend.title = element_blank(),
    legend.justification = "bottom",
    legend.position = c(0.5, 0))

travel_times_dir1 %>%
  ggplot(aes(time, travel_diff)) +
  geom_line() +
  facet_wrap(vars(day)) +
  ggtitle(directions[1]) +
  theme_bw()

# Direction 2
travel_times_dir2 <- travel_times %>% 
  filter(direction == directions[2])

travel_times_piv_dir2 <- travel_times_piv %>% 
  filter(direction == directions[2])

travel_times_piv_dir2 %>%
  ggplot(aes(time, travel_time, color = type)) +
  geom_line() +
  facet_wrap(vars(day)) +
  ggtitle(directions[2]) +
  theme_bw() +
  theme(
  legend.title = element_blank(),
  legend.justification = "bottom",
  legend.position = c(0.5, 0))

travel_times_dir2 %>%
  ggplot(aes(time, travel_diff)) +
  geom_line() +
  facet_wrap(vars(day)) +
  ggtitle(directions[2]) +
  theme_bw()

# Scatter plots
travel_times %>% 
  ggplot(aes(bus_time, links_time)) +
  geom_point(alpha = 0.3) +
  geom_smooth(se = FALSE) +
  theme_bw()

travel_times %>% 
  ggplot(aes(bus_time, links_time)) +
  geom_point(alpha = 0.3) +
  geom_smooth(se = FALSE) +
  facet_wrap(vars(direction), scales = "free") +
  theme_bw()
```

### Correlation between bus travel time and links travel time

```{r}
paste("Correlation:", round(with(travel_times, cor(bus_time, links_time)), 2))

travel_times %>% 
  group_by(direction) %>% 
  summarise(correlation = round(cor(bus_time, links_time), 2))
```

## Average pattern of links time and bus trips time

```{r}
# Average total links time in each 30 minute bin
links_30min <- link_stats %>%
  group_by(logtime, direction) %>%
  summarise(links_time = sum(tt)) %>%
  ungroup() %>% 
  mutate(logtime = ceiling_date(logtime, "30 mins"),
         time = as_hms(logtime)) %>% 
  group_by(time, direction) %>% 
  summarise(links_time = mean(links_time)) %>% 
  ungroup() %>% 
  mutate(rush = case_when(
           between(time, as_hms("06:00:00"), as_hms("10:00:00")) ~ "morning",
           between(time, as_hms("15:00:00"), as_hms("19:00:00")) ~ "evening",
           TRUE ~ "neither"),
         rush = factor(rush, levels = c("morning", "evening", "neither"))) %>% 
  relocate(links_time, .after = everything())

# Average bus trip times in each 30 minute bin
trip_times_30min <- trip_times %>% 
  mutate(arrival_time = ceiling_date(arrival_time, "30 minutes"),
         time = as_hms(arrival_time)) %>% 
  group_by(time, direction) %>% 
  summarise(bus_time = round(mean(trip_time))) %>%
  ungroup() %>% 
  mutate(rush = case_when(
           between(time, as_hms("06:00:00"), as_hms("10:00:00")) ~ "morning",
           between(time, as_hms("15:00:00"), as_hms("19:00:00")) ~ "evening",
           TRUE ~ "neither"),
         rush = factor(rush, levels = c("morning", "evening", "neither"))) %>% 
  relocate(bus_time, .after = everything())

links_trips_30min <- links_30min %>% 
  inner_join(trip_times_30min) %>% 
  pivot_longer(cols = links_time:bus_time, names_to = "type", values_to = "value")

ggplot(links_trips_30min, aes(time, value)) +
  geom_line(alpha = 0.2) +
  geom_line(data = filter(links_trips_30min, rush == "morning"), color = "blue", size = 0.5) +
  geom_line(data = filter(links_trips_30min, rush == "evening"), color = "blue", size = 0.5) +
  facet_grid(rows = vars(type), cols = vars(direction), scales = "free_y") +
  annotate("rect", xmin = as_hms("06:00:00"), xmax = as_hms("10:00:00"), ymin = -Inf, ymax = Inf,
           fill = "red", alpha = .1) +
  annotate("rect", xmin = as_hms("15:00:00"), xmax = as_hms("19:00:00"), ymin = -Inf, ymax = Inf,
           fill = "red", alpha = .1) +
  scale_x_time(limits = as_hms(c("06:00:00", "23:59:59")),
               breaks = as_hms(c("06:00:00", "08:00:00", "10:00:00",
                                 "15:00:00", "17:00:00", "19:00:00")),
               labels = c("6", "8", "10", "15", "17", "19")) +
  labs(x = "Time", y = "") +
  theme_bw()
```

## Correlations

```{r}
standardiser <- function(basis = hour) {
  if (!deparse(substitute(basis)) %in% c("time", "hour", "rush")) {
    stop("basis must be either time, hour, or rush")
  }
  
  trip_times_std <- trip_times_agg %>%
    group_by({{basis}}, direction) %>% 
    mutate(
      period_buses_no = n(),
      bus_time_std = as.vector(scale(bus_time))) %>% 
    ungroup() %>% 
    mutate(bus_time_std = if_else(is.nan(bus_time_std), 0, bus_time_std)) %>%
    filter(abs(bus_time_std) < 3) %>%
    select(-bus_time)
    
  links_times_std <- links_times %>% 
    group_by({{basis}}, direction) %>% 
    mutate(links_time_std = as.vector(scale(links_time))) %>% 
    ungroup() %>% 
    filter(abs(links_time_std) < 3) %>%
    select(-links_time)
  
  # Combine
  travel_times_std <<- trip_times_std %>%
    inner_join(links_times_std) %>%
    mutate(std_travel_diff = bus_time_std - links_time_std)
  
  travel_times_std_piv <<- travel_times_std %>% 
    pivot_longer(cols = bus_time_std:links_time_std, names_to = "type", values_to = "std_travel_time")
}

standardiser()
```

#### Visualise bus time and links time

```{r}
travel_times_std %>% 
  ggplot(aes(bus_time_std)) +
  geom_histogram() +
  theme_bw()

travel_times_std %>% 
  filter(rush != "neither") %>% 
  ggplot(aes(bus_time_std)) +
  geom_histogram() +
  facet_grid(vars(rush), vars(direction), scales = "free_y") +
  theme_bw()

travel_times_std %>% 
  ggplot(aes(bus_time_std)) +
  geom_boxplot() +
  theme_bw()

travel_times_std %>% 
  filter(rush != "neither") %>% 
  ggplot(aes(bus_time_std)) +
  geom_boxplot() +
  facet_grid(vars(rush), vars(direction)) +
  theme_bw()

travel_times_std %>% 
  ggplot(aes(links_time_std)) +
  geom_histogram() +
  theme_bw()

travel_times_std %>% 
  filter(rush != "neither") %>% 
  ggplot(aes(links_time_std)) +
  geom_histogram() +
  facet_grid(vars(rush), vars(direction), scales = "free_y") +
  theme_bw()

travel_times_std %>% 
  ggplot(aes(links_time_std)) +
  geom_boxplot() +
  theme_bw()

travel_times_std %>% 
  filter(rush != "neither") %>% 
  ggplot(aes(links_time_std)) +
  geom_boxplot() +
  facet_grid(vars(rush), vars(direction)) +
  theme_bw()
```

```{r}
# Morning Northbound
travel_times_std_NB_morn <- travel_times_std %>% 
  filter(direction == "NB",
         rush == "morning")

travel_times_std_piv_NB_morn <- travel_times_std_piv %>% 
  filter(direction == "NB",
         rush == "morning")

travel_times_std_piv_NB_morn %>%
  ggplot(aes(time, std_travel_time, color = type)) +
  geom_line() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  ylim(c(-3,3)) +
  facet_wrap(vars(day)) +
  ggtitle("Northbound") +
  theme_bw() +
  theme(
    legend.title = element_blank(),
    legend.justification = "bottom",
    legend.position = c(0.5, 0))

travel_times_std_NB_morn %>%
  ggplot(aes(time, std_travel_diff)) +
  geom_line() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(vars(day)) +
  ggtitle("Northbound") +
  theme_bw()

# Evening Southbound
travel_times_std_SB_eve <- travel_times_std %>% 
  filter(direction == "SB",
         rush == "evening")

travel_times_std_piv_SB_eve <- travel_times_std_piv %>% 
  filter(direction == "SB",
         rush == "evening")

travel_times_std_piv_SB_eve %>%
  ggplot(aes(time, std_travel_time, color = type)) +
  geom_line() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  ylim(c(-3,3)) +
  facet_wrap(vars(day)) +
  ggtitle("Southbound") +
  theme_bw() +
  theme(
    legend.title = element_blank(),
    legend.justification = "bottom",
    legend.position = c(0.5, 0))

travel_times_std_SB_eve %>%
  ggplot(aes(time, std_travel_diff)) +
  geom_line() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(vars(day)) +
  ggtitle("Southbound") +
  theme_bw()
```

#### Scatter plots

```{r}
travel_times_std %>% 
  ggplot(aes(links_time_std, bus_time_std)) +
  geom_point(alpha = 0.3) +
  geom_smooth(se = FALSE) +
  ggtitle("All Times") +
  theme_bw()

travel_times_std %>% 
  filter(rush != "neither") %>% 
  ggplot(aes(links_time_std, bus_time_std)) +
  geom_point(alpha = 0.3) +
  geom_smooth(se = FALSE) +
  ggtitle("Peak Times") +
  theme_bw()

travel_times_std %>% 
  filter(rush != "neither") %>% 
  ggplot(aes(links_time_std, bus_time_std)) +
  geom_point(alpha = 0.3) +
  geom_smooth(se = FALSE) +
  facet_grid(vars(rush), vars(direction)) +
  theme_bw()
```

#### Correlation coefficients

```{r}
paste("All times correlation:", round(with(travel_times_std, cor(bus_time_std, links_time_std)), 2))

paste("Peak times correlation:", round(with(filter(travel_times_std, rush != "neither"), 
                                            cor(bus_time_std, links_time_std)), 2))

# Peak times and direction correlation
travel_times_std %>% 
  filter(rush != "neither") %>% 
  group_by(rush, direction) %>% 
  summarise(correlation = round(cor(bus_time_std, links_time_std), 2)) %>% 
  ungroup()

# Time period correlation
travel_times_std_cor <- travel_times_std %>%
  filter(rush != "neither") %>%
  group_by(time, rush, direction) %>% 
  summarise(
    correlation = cor(bus_time_std, links_time_std),
    sample_size = n()) %>% 
  ungroup() %>% 
  na.omit()

travel_times_std_cor %>% 
  ggplot(aes(time, correlation)) +
  geom_line() +
  facet_grid(vars(direction), vars(rush), scales = "free_x") +
  theme_bw()

# Hourly correlation
travel_times_std_hr_cor <- travel_times_std %>%
  filter(rush != "neither") %>%
  group_by(hour, rush, direction) %>% 
  summarise(
    correlation = cor(bus_time_std, links_time_std),
    sample_size = n()) %>% 
  ungroup() %>% 
  na.omit()

travel_times_std_hr_cor %>% 
  ggplot(aes(hour, correlation, label = round(correlation, 2))) +
  geom_line() +
  geom_label() +
  facet_grid(vars(direction), vars(rush), scales = "free_x") +
  theme_bw()
```

#### Linear Regression

##### All times

```{r}
# Try all variables
all_lm <- lm(bus_time_std ~ links_time_std + rush*direction, travel_times_std)
summary(all_lm)
drop1(all_lm)
step(all_lm, trace = FALSE)

# Only links_time_std is needed
all_lm <- lm(bus_time_std ~ links_time_std, travel_times_std)
summary(all_lm)
tidy(all_lm)
glance(all_lm)

# Residual diagnostics
augment(all_lm) %>% 
  ggplot(aes(.fitted, .resid)) +
  geom_point() +
  geom_smooth(se = FALSE) +
  theme_bw()

augment(all_lm) %>% 
  ggplot(aes(.resid)) +
  geom_histogram() +
  theme_bw()

all_lm %>%
  ggplot(aes(sample=.resid)) +
  geom_qq() + 
  geom_qq_line() +
  theme_bw()
```

##### Peak Times

```{r}
# Morning Northbound
mor_NB_lm <- lm(bus_time_std ~ links_time_std, travel_times_std_NB_morn)
summary(mor_NB_lm)
tidy(mor_NB_lm)
glance(mor_NB_lm)

# Residual diagnostics
augment(mor_NB_lm) %>% 
  ggplot(aes(.fitted, .resid)) +
  geom_point() +
  geom_smooth(se = FALSE) +
  theme_bw()

augment(mor_NB_lm) %>% 
  ggplot(aes(.resid)) +
  geom_histogram() +
  theme_bw()

mor_NB_lm %>%
  ggplot(aes(sample=.resid)) +
  geom_qq() + 
  geom_qq_line() +
  theme_bw()

# Evening Southbound
eve_SB_lm <- lm(bus_time_std ~ links_time_std, travel_times_std_SB_eve)
summary(eve_SB_lm)
tidy(eve_SB_lm)
glance(eve_SB_lm)

# Residual diagnostics
augment(eve_SB_lm) %>% 
  ggplot(aes(.fitted, .resid)) +
  geom_point() +
  geom_smooth(se = FALSE) +
  theme_bw()

augment(eve_SB_lm) %>% 
  ggplot(aes(.resid)) +
  geom_histogram() +
  theme_bw()

eve_SB_lm %>%
  ggplot(aes(sample=.resid)) +
  geom_qq() + 
  geom_qq_line() +
  theme_bw()
```
