---
title: "Initial Exploration"
author: "Ibrahim Al-Hindi"
date: "2022-08-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Libraries
library(tidyverse)
library(reticulate)
library(skimr)
library(lubridate)
library(ggmap)
library(geosphere)
library(sf)
library(geodist)
library(dodgr)
```

```{r}
# Read in data
probe1 <- read_csv("Data/SingleProbe.csv")
probe2 <- read_csv("Data/SingleProbe2.csv")
```



# Data Exploration

## Probes

```{r}
probe2 <- probe2 %>% 
  mutate(ingest_hour...11 = as.numeric(ingest_hour...11))

tot_probes <- probe1 %>% 
  bind_rows(probe2)

probes <- tot_probes %>%
  # remove duplicates if the variables below are identical
  distinct(logtime, siteid, probeid, firstseenat, id, number, name, beaconindex,
           instance_id, androidname, longitude, latitude,
           .keep_all = TRUE) %>% 
  mutate(logtime = as_datetime(str_remove(logtime, " \\+10:30"), tz = "UTC"),
         firstseenat = as_datetime(str_remove(firstseenat, " \\+10:30"), tz = "UTC")) %>% 
  arrange(probeid, logtime)

# different information(s) for the same logtime, probeid, and firsseenat
discrepencies <- probes %>% 
  group_by(logtime, probeid, firstseenat) %>% 
  filter(
      n_distinct(siteid) > 1 |
      n_distinct(id) > 1 |
      n_distinct(number) > 1 |
      n_distinct(name) > 1 |
      n_distinct(beaconindex) > 1 |
      n_distinct(instance_id) > 1 |
      n_distinct(androidname) > 1 |
      n_distinct(longitude) > 1 |
      n_distinct(latitude) > 1) %>% 
  ungroup() %>% 
  arrange(logtime)
```

```{r}
probes %>% 
  count(probeid)
```


### Calculate Speeds

#### Calculate Distances

Distance is calculated using haversine distance, therefore does not consider roads.
gmapdistance package can calculate driving distance using Google Maps but Google API used needs payment

```{r}
# need to create a for loop because the geodist function does not work with grouped data
df <- tibble(cols = names(probes),
               values = NA) %>%
    pivot_wider(names_from = cols, values_from = values) %>%
    # rename because geodist only recognises "lat" and "lon"
    rename(lat = latitude, lon = longitude)

for (probe in unique(probes$probeid)) {
    df1 <- probes %>% 
    filter(probeid == probe) %>% 
    rename(lat = latitude, lon = longitude)
  
  distances <- df1 %>%  
    geodist(sequential = TRUE, measure = "haversine")
  
  df1 <- df1 %>% 
    mutate(distance = c(0, distances))
  
  df <- df %>% 
    bind_rows(df1)
}

df <- df[-1,] %>% 
  relocate(distance, .after = lat) %>% 
  rename(latitude = lat, longitude = lon)

# convert meters to KM
probes <- df %>% 
  mutate(distance = distance/1000)
```

```{r}
ggplot(probes, aes(distance)) +
  geom_histogram()

ggplot(probes, aes(distance)) +
  geom_histogram() +
  facet_wrap(vars(probeid))
```
probe 260 problematic

#### Calculate Times

```{r}
probes <- probes %>%
  group_by(probeid) %>% 
  mutate(time = as.numeric(difftime(logtime, lag(logtime)), units = "hours"),
         time = replace_na(time, 0)) %>% 
  relocate(time, .after = logtime) %>% 
  ungroup()
```

```{r}
ggplot(probes, aes(time)) +
  geom_histogram()

ggplot(probes, aes(time)) +
  geom_histogram() +
  facet_wrap(vars(probeid))
```

looks fine


```{r}
ggplot(probes, aes(distance, time)) +
  geom_point()

ggplot(probes, aes(distance, time)) +
  geom_point() +
  facet_wrap(vars(probeid))
```

probe 260 looks very problematic, a few points for 315 looks problematic

#### Calculate Speeds

```{r}
probes <- probes %>% 
  mutate(speed = distance/time)
```

```{r}
ggplot(probes, aes(speed)) +
  geom_histogram()

ggplot(probes, aes(speed)) +
  geom_histogram() +
  facet_wrap(vars(probeid), scales = "free")
```



```{r}
high_speeds <- probes %>% 
  filter(speed > 200) %>% 
  select(probeid, time, distance, speed)

ggplot(high_speeds, aes(x = as.character(probeid))) +
  geom_bar()
```

probeid 968243260 looks especially problematic

#### Average Speeds Between Each Pair of Beacons

possible exclude the following sites first

```{r}
multi_coords <- probes %>% 
  group_by(siteid) %>% 
  summarise(nums = n_distinct(latitude, longitude)) %>% 
  arrange(-nums) %>% 
  ungroup() %>% 
  filter(nums > 1) %>% 
  pull(siteid)

multi_coords_probes <- probes %>% 
  filter(siteid %in% multi_coords) %>% 
  arrange(siteid, name)
```

several sites have more than one pair of coordinates





















# Map

```{r}
adel_map <- get_map(c(left = min(probes$longitude) - 0.005,
                      bottom = min(probes$latitude) - 0.005,
                      right = max(probes$longitude) + 0.005,
                      top = max(probes$latitude)) + 0.005,
                    source = "stamen")

ggmap(adel_map) +
  geom_point(data = probes, aes(x = longitude, y = latitude), color = "red")





qmplot(longitude, latitude, data = probes, maptype = "toner-lite", color = I("red"))
  


```







