```{r libraries}
# Libraries
library(tidyverse)
library(here)
library(lubridate)
library(leaflet)
library(hms)
library(janitor)
library(broom)
```

```{r data}
# Data

## Holidays dates
holidays_raw <- read_csv(here("data/DIT/specialdays.csv"))

## Links information
links_raw <- read_csv(here("data/DIT/links.csv"))

## List of links on segment and their geometries
links_segment_raw <- read_csv(here("data/South-Rd/Links - South Rd.csv"))

## Stops information
full_stops_raw <- read_csv(here("data/GTFS/stops.csv"))

## Routes, trips, and stops on segment
segment_route_stops_raw <- read_csv(here("data/South-Rd/South Rd Routes and Trips gtfs 1157.csv"))

## GTFS-R Trip Updates
updates_raw <- read_csv(here("data/South-Rd/trip_updates_SouthRd_Mar.csv"))

## Link 5 minute statistics
link_stats_raw <- read_csv(here("data/South-Rd/link_stats_Mar.csv"))
```

# Data Wrangling and Exploration

## Holidays

```{r}
holidays <- unique(holidays_raw$date)
rm(holidays_raw)
```

## Links Segment

```{r}
# Limit updates to end of period examined and get most recent update per link,
# this is done to get the most recent enabled status for each link
links <- links_raw %>%
  filter(date(dms_update_ts) < "2022-04-01") %>%
  group_by(id) %>%
  filter(dms_update_ts == max(dms_update_ts)) %>%
  ungroup() %>%
  select(id, originid:length, direction) %>%
  distinct()

links_segment <- links_segment_raw %>%
  rename_with(tolower) %>%
  mutate(
    latitude = round(latitude, 5),
    longitude = round(longitude, 5),
    # Extract name of locations where link starts and ends
    start_loc = str_extract(name, "(?<=- ).*(?= to)"),
    end_loc = str_extract(name, "(?<=to ).*")) %>%
  # Extract the latitudes of where the link starts and ends
  group_by(linkid) %>%
  mutate(
    start_lat = first(latitude, order_by = ordernumber),
    end_lat = last(latitude, order_by = ordernumber)) %>%
  ungroup() %>%
  # Length in links_segment_raw is not correct, join with links to get correct lengths and direction
  select(-length) %>%
  left_join(links, by = c("linkid" = "id"))

# Get only first entry per link
links_segment_unq <- links_segment %>%
  arrange(linkid, ordernumber) %>%
  distinct(linkid, .keep_all = TRUE)

directions <- sort(unique(links_segment_unq$direction))

rm(links_raw)
rm(links_segment_raw)
```

## Full Stops and Route-Trips-Stops

```{r}
segment_route_stops <- clean_names(segment_route_stops_raw)
  
segment_routes <- segment_route_stops %>% 
  pull(route_id) %>% 
  unique()

segment_stops <- segment_route_stops %>% 
  pull(stop_id_gtfs_history_prod_stops_csv) %>% 
  unique()

# Join stops information to stops on the segment
stops_segment <- full_stops_raw %>%
  filter(stop_id %in% segment_stops) %>%
  select(stop_id, stop_name, stop_desc, stop_lat, stop_lon) %>%
  distinct(stop_id, .keep_all = TRUE) %>%
  # If stop is on East side, that means trip SB, and vice versa
  mutate(direction = if_else(str_detect(stop_name, "East"), "SB", "NB"))

rm(segment_route_stops_raw)
rm(full_stops_raw)
```

## Updates

```{r}
# Keep only updates for stops on segment and remove weekends and public holidays
updates <- updates_raw %>%
  rename(
    trip_id = id,
    vehicle_id = label) %>%
  mutate(
    start_date = ymd(start_date),
    timestamp = as_datetime(timestamp, tz = "Australia/Adelaide"),
    arrival_time = as_datetime(arrival_time, tz = "Australia/Adelaide")) %>%
  select(-direction_id, -wheelchair_accessible) %>% 
  inner_join(select(stops_segment, stop_id, direction), by = "stop_id") %>%
  filter(
    !start_date %in% holidays,
    !wday(start_date, label = TRUE) %in% c("Sat", "Sun"))

# Clean updates and get time to each stop for error detection
updates_stop_times <- updates %>% 
  group_by(start_date, trip_id) %>%
  arrange(stop_sequence, .by_group = TRUE) %>%
  # Remove trips where all the stops have a delay over 2400 (40 minutes) or
  # early by more than 900 (15 minutes), ie entire trip is delayed or early.
  # Most likely error due to entering the information later. This is done to prevent
  # incorrect analysis since they will be in the wrong time period
  filter(!(all(delay > 2400) |
           all(delay < -900))) %>%
  # Remove one-off large jumps in between two observations
  filter(!(delay - lag(delay, order_by = stop_sequence, default = 0) > 1000 &
             delay - lead(delay, order_by = stop_sequence, default = 0) > 1000)) %>%
  # Remove any observations that have an arrival_time later than any following
  # arrival_time in the trip AND the timestamp is earlier than any following timestamps in the trip.
  # This ensures the most recent timestamp is preferred when discrepancy occurs
  filter(!(as.numeric(arrival_time) > order_by(-stop_sequence, cummin(as.numeric(arrival_time))) &
             as.numeric(timestamp) == order_by(-stop_sequence, cummin(as.numeric(timestamp))))) %>%
  # Remove if the arrival_time is less than the previous arrival_times AND the timestamp
  # is older than the previous timestamps
  filter(!(arrival_time < order_by(stop_sequence, cummax(as.numeric(arrival_time))) &
             timestamp < order_by(stop_sequence, cummax(as.numeric(timestamp))))) %>%
  # If arrival_time of a stop is less than prior stops but they all have the same
  # timestamp, it's not possible to know which is correct. Assume earlier stop_sequence
  # is correct since it is closer when the update is made
  filter(!(timestamp == order_by(stop_sequence, cummax(as.numeric(timestamp))) &
             arrival_time < order_by(stop_sequence, cummax(as.numeric(arrival_time))))) %>%
  # If after the above, two consecutive stops have the same arrival_time, remove
  # the one with an older timestamp
  filter(!(arrival_time == lead(arrival_time, 
                                order_by = stop_sequence,
                                default = ymd("9999-01-01")) &
             timestamp < lead(timestamp, 
                              order_by = stop_sequence,
                              default = ymd("9999-01-01")))) %>%
  # If two stops have the same arrival_time and same time_stamp, remove the one
  # with a higher stop sequence. Make sure only 2 stops and not more
  filter(!(arrival_time != lead(arrival_time, 
                                order_by = stop_sequence,
                                default = ymd("9999-01-01")) &
             arrival_time == lag(arrival_time, 
                                 order_by = stop_sequence,
                                 default = ymd("0000-01-01")) &
             timestamp == lag(timestamp, 
                              order_by = stop_sequence,
                              default = ymd("0000-01-01")))) %>%
  # Remove trips with multiple repeating arrival_times
  filter(!n_distinct(arrival_time) < n()) %>%
  mutate(
    to_stop_time = as.numeric(arrival_time - lag(arrival_time), units = "secs"),
    first_stop = first(stop_id, order_by = stop_sequence),
    last_stop = last(stop_id, order_by = stop_sequence)) %>% 
  ungroup() %>%
  mutate(to_stop_time = replace_na(to_stop_time, 0))

paste0("Percentage of errors: ", round((nrow(updates) - nrow(updates_stop_times)) / nrow(updates) * 100, 2), "%")

# Take top occurring stops pair for each direction
top_first_last_stops <- updates_stop_times %>% 
  distinct(start_date, trip_id, direction, first_stop, last_stop) %>%
  group_by(direction) %>% 
  count(first_stop, last_stop, sort = TRUE) %>% 
  slice_max(order_by = n) %>% 
  ungroup()

# Get trips with only most occurring stops pair per direction identified
updates_stop_times <- updates_stop_times %>%
  semi_join(top_first_last_stops, by = c("direction", "first_stop", "last_stop"))

# Get total time per trip within segment
updates_trip_time <- updates_stop_times %>% 
  group_by(start_date, trip_id) %>%
  arrange(stop_sequence, .by_group = TRUE) %>%
  mutate(
    trip_time = as.numeric((last(arrival_time) -first(arrival_time)), units = "secs"),
    delay_diff = abs(last(delay) - first(delay))) %>% 
  ungroup() %>%
  # If the delay_diff is greater than 600 (10 minutes), most likely an error. Remove
  filter(delay_diff < 600) %>% 
  arrange(start_date, trip_id, stop_sequence)


# For further analysis, take first stop only from each trip because the arrival_time
# of the first stop will be considered as the trip start time used as basis for aggregation later.
# Also add whether each trip occurrs during either of the peak times
trip_times <- updates_trip_time %>%
  group_by(start_date, trip_id) %>%
  filter(stop_sequence == first(stop_sequence, order_by = stop_sequence)) %>%
  ungroup() %>%
  distinct(start_date, trip_id, .keep_all = TRUE) %>% 
  mutate(
    rush = case_when(
      as_hms("06:30:00") <= as_hms(arrival_time) & as_hms(arrival_time) < as_hms("10:00:00") ~ "Morning",
      as_hms("15:30:00") <= as_hms(arrival_time) & as_hms(arrival_time) < as_hms("19:00:00") ~ "Evening",
      TRUE ~ "Neither"),
    rush = factor(rush, levels = c("Morning", "Evening", "Neither"))) %>% 
  select(-delay_diff)

rm(updates_raw)
```

```{r}
# Split trips into 5 minute intervals average the trips times per interval
trip_times_agg <- trip_times %>% 
  mutate(arrival_time = ceiling_date(arrival_time, "5 minutes")) %>% 
  group_by(arrival_time, direction) %>% 
  summarise(
    number_buses = n(),
    bus_time = round(mean(trip_time))) %>%
  ungroup() %>% 
  mutate(
    day = date(arrival_time),
    time = as_hms(arrival_time),
    hour = hour(time),
    rush = case_when(
      between(time, as_hms("06:30:00"), as_hms("09:59:00")) ~ "Morning",
      between(time, as_hms("15:30:00"), as_hms("18:59:00")) ~ "Evening",
      TRUE ~ "Neither"),
    rush = factor(rush, levels = c("Morning", "Evening", "Neither"))) %>% 
  relocate(day:rush, .after = arrival_time) %>% 
  select(-arrival_time)
```

# Getting Information for Analysis

## Identify Routes

Routes were identified by examining the [network map](https://www.adelaidemetro.com.au/__data/assets/pdf_file/0009/824247/Adelaide-Metro-network-map.pdf).

All the routes were overlayed on a map on Tableau and routes on the segment were selected and exported.

## Identify Stops

Same as above

## Identify Links

### Direction 1

```{r}
# The going back and forth between links_dir process is for the purpose of
# producing links_disabled 

links_dir1 <- links_segment_unq %>%
  # Build map from south to north
  filter(direction == directions[1]) %>% 
  arrange(start_lat)

links_dir1_enabled <- links_dir1 %>%
  filter(enabled)

# Start and end at same locations as bus stops by referencing map above
dir1_start_loc <- "celtic"
dir1_end_loc <- "anzac"

# Retrieve sequence of links composing segment
links_dir1_enabled <- links_dir1_enabled[min(which(str_starts(tolower(links_dir1_enabled$start_loc),
                                                                  tolower(dir1_start_loc)))):
                                               max(which(str_starts(tolower(links_dir1_enabled$end_loc),
                                                                    tolower(dir1_end_loc)))),] %>% 
  group_by(start_loc) %>%
  filter(length == min(length)) %>%
  ungroup()

# Obtain shape for each link
links_dir1_map <- links_segment %>% 
  semi_join(links_dir1_enabled, by = "linkid") %>% 
  arrange(start_lat, ordernumber)

# Create map

links_colors <- c("green","red")

n <- 1

map_dir1 <- leaflet() %>% 
  addTiles()

# Iterate over links to add polyline of each link to the map
for (link in unique(links_dir1_map$linkid)) {
  map_df <- links_dir1_map %>% 
    filter(linkid == link)
  
  map_dir1 <- map_dir1 %>% 
    addPolylines(data = map_df, lng = ~longitude, lat = ~latitude, color = links_colors[n],
                 label = ~linkid, opacity = 0.3)
  
  # Switch colors
  if (n == 1) {
    n <- 2
  } else {
    n <- 1
  }
}

dir1_stops <- stops_segment %>%
  filter(stop_id %in% c(top_first_last_stops[top_first_last_stops["direction"] == directions[1],]$first_stop,
                        top_first_last_stops[top_first_last_stops["direction"] == directions[1],]$last_stop))

map_dir1 %>% 
  addCircles(data = dir1_stops, lat = ~stop_lat, lng = ~stop_lon, label = ~ stop_id, radius = 50, opacity = 1)
```

### Direction 2

```{r}
# Same process as direction 1

links_dir2 <- links_segment_unq %>%
  filter(direction == directions[2]) %>% 
  arrange(-start_lat)

links_dir2_enabled <- links_dir2 %>%
  filter(enabled)

dir2_start_loc <- "anzac"
dir2_end_loc <- "celtic"

links_dir2_enabled <- links_dir2_enabled[min(which(str_starts(tolower(links_dir2_enabled$start_loc),
                                                                  tolower(dir2_start_loc)))):
                                               max(which(str_starts(tolower(links_dir2_enabled$end_loc),
                                                                    tolower(dir2_end_loc)))),] %>% 
  group_by(start_loc) %>%
  filter(length == min(length)) %>%
  ungroup()

links_dir2_map <- links_segment %>% 
  semi_join(links_dir2_enabled, by = "linkid") %>% 
  arrange(-start_lat, ordernumber)

links_colors <- c("green","red")

n <- 1

map_dir2 <- leaflet() %>% 
  addTiles()

for (link in unique(links_dir2_map$linkid)) {
  map_df <- links_dir2_map %>% 
    filter(linkid == link)
  
  map_dir2 <- map_dir2 %>% 
    addPolylines(data = map_df, lng = ~longitude, lat = ~latitude, color = links_colors[n],
                 label = ~linkid, opacity = 0.3)

  if (n == 1) {
    n <- 2
  } else {
    n <- 1
  }
}

dir2_stops <- stops_segment %>%
  filter(stop_id %in% c(top_first_last_stops[top_first_last_stops["direction"] == directions[2],]$first_stop,
                        top_first_last_stops[top_first_last_stops["direction"] == directions[2],]$last_stop))

map_dir2 %>% 
  addCircles(data = dir2_stops, lat = ~stop_lat, lng = ~stop_lon, label = ~ stop_id, radius = 50, opacity = 1)
```

#### Both directions together

```{r}
# Plot links from both directions
map_both <- leaflet() %>% 
  addTiles()

# Add links first direction
for (link in unique(links_dir1_map$linkid)) {
  map_df <- links_dir1_map %>% 
    filter(linkid == link)
  
  map_both <- map_both %>% 
    addPolylines(data = map_df, lng = ~longitude, lat = ~latitude, color = "green", 
                 opacity = 0.3, label = ~linkid)
}

# Add links second direction
for (link in unique(links_dir2_map$linkid)) {
  map_df <- links_dir2_map %>% 
    filter(linkid == link)
  
  map_both <- map_both %>% 
    addPolylines(data = map_df, lng = ~longitude, lat = ~latitude, color = "red",
                 opacity = 0.3, label = ~linkid)
}

map_both %>% 
  addCircles(data = dir1_stops, lat = ~stop_lat, lng = ~stop_lon, label = ~ stop_id, radius = 50, opacity = 1) %>%
  addCircles(data = dir2_stops, lat = ~stop_lat, lng = ~stop_lon, label = ~ stop_id, radius = 50, opacity = 1)
```

### Missing Links

```{r}
# DF for missing links imputation
missing_links_impute <- tibble(from = character(),
                               to = character(),
                               linkid = numeric(),
                               prior_linkid = numeric(),
                               post_linkid = numeric(),
                               direction = character())
```

#### Direction 1 Missing Links

```{r}
# For the missing link, retrieve the link as well closest enabled previous and 
# proceeding links for imputation later
missing_links_dir1 <- tibble(from = character(),
                        to = character())

# Add start and end locations of missing links
missing_links_dir1 <- missing_links_dir1 %>% 
  add_row(from = "walsh",
          to = "celtic")

for (i in 1:nrow(missing_links_dir1)) {
  
  from <- tolower(missing_links_dir1$from[i])
  to <- tolower(missing_links_dir1$to[i])
  
  for (a in 1:nrow(links_dir1)) {
    # Get linkid for missing link
    if (str_starts(tolower(links_dir1$start_loc[a]), from) && str_starts(tolower(links_dir1$end_loc[a]), to)) {
      
      linkid <- links_dir1$linkid[a]

      # Retrieve last prior link that is enabled
      prior_linkid <- links_dir1[1:a-1,] %>%
        filter(enabled) %>%
        slice_tail() %>%
        pull(linkid)

      # Retrieve first proceeding link that is enabled
      post_linkid <- links_dir1[a+1:nrow(links_dir1),] %>%
        filter(enabled) %>%
        slice_head() %>%
        pull(linkid)

      missing_links_impute <- missing_links_impute %>%
        add_row(from = from,
                to = to,
                linkid = linkid,
                prior_linkid = prior_linkid,
                post_linkid = post_linkid,
                direction = directions[1])
      break
    }
  }
}
```

#### Direction 2 Missing Links

```{r}
missing_links_dir2 <- tibble(from = character(),
                        to = character())

missing_links_dir2 <- missing_links_dir2 %>% 
  add_row(from = "celtic",
          to = "walsh")

for (i in 1:nrow(missing_links_dir2)) {
  
  from <- tolower(missing_links_dir2$from[i])
  to <- tolower(missing_links_dir2$to[i])
  
  for (a in 1:nrow(links_dir2)) {
    if (str_starts(tolower(links_dir2$start_loc[a]), from) && str_starts(tolower(links_dir2$end_loc[a]), to)) {
      
      linkid <- links_dir2$linkid[a]
      
      prior_linkid <- links_dir2[1:a-1,] %>%
        filter(enabled) %>%
        slice_tail() %>%
        pull(linkid)

      post_linkid <- links_dir2[a+1:nrow(links_dir1),] %>%
        filter(enabled) %>%
        slice_head() %>%
        pull(linkid)

      missing_links_impute <- missing_links_impute %>%
        add_row(from = from,
                to = to,
                linkid = linkid,
                prior_linkid = prior_linkid,
                post_linkid = post_linkid,
                direction = directions[2])
      break
    }
  }
}
```

```{r}
# Plot links from both directions
map_both <- leaflet() %>% 
  addTiles()

# Add links first direction
for (link in unique(links_dir1_map$linkid)) {
  map_df <- links_dir1_map %>% 
    filter(linkid == link)
  
  map_both <- map_both %>% 
    addPolylines(data = map_df, lng = ~longitude, lat = ~latitude, color = "green", 
                 opacity = 0.3, label = ~linkid)
}

# Add links second direction
for (link in unique(links_dir2_map$linkid)) {
  map_df <- links_dir2_map %>% 
    filter(linkid == link)
  
  map_both <- map_both %>% 
    addPolylines(data = map_df, lng = ~longitude, lat = ~latitude, color = "red",
                 opacity = 0.3, label = ~linkid)
}

map_both

# Save linkid's for both directions, the missing links, and extra links needed to impute missing links
links_all <- unique(c(links_dir1_enabled$linkid, links_dir2_enabled$linkid,
                missing_links_impute$linkid,missing_links_impute$prior_linkid, 
                missing_links_impute$post_linkid))
```

### Link Stats

Links stats are retrieved from AWS Athena database using the links identified to construct the segments and the links needed to impute missing links stats

```{r}
link_stats <- link_stats_raw %>% 
  select(logtime:avgdensity) %>%
  mutate(logtime = ymd_hms(str_remove(logtime, " \\+10:30"), tz = "Australia/Adelaide")) %>%
  filter(
    !as_date(logtime) %in% holidays,
    !wday(logtime, label = TRUE) %in% c("Sat", "Sun")) %>% 
  distinct()

# Join link_stats with links_segment to get direction and length
link_stats <- link_stats %>%
  left_join(select(links_segment_unq, linkid, direction, length)) %>%
  # Calculate speed km/h
  mutate(speed = round((length / 1000) / (tt / 3600))) %>% 
  # Remove speeds over 150 km/h as most likely errors
  filter(speed < 150)

summary(link_stats)

rm(link_stats_raw)
```

### Check for links with no stats

```{r}
links_with_stats <- links_segment %>% 
  filter(linkid %in% c(links_dir1_enabled$linkid, links_dir2_enabled$linkid,
                       missing_links_impute$linkid)) %>% 
  semi_join(link_stats, by = "linkid")

links_no_stats <- links_segment %>% 
  filter(linkid %in% missing_links_impute$linkid) %>%  
  anti_join(link_stats, by = "linkid")
  
map_links_with_stats <- leaflet() %>% 
  addTiles()

# Draw links with stats in green
for (link in unique(links_with_stats$linkid)) {
  map_df <- links_with_stats %>% 
    filter(linkid == link)
  
  map_links_with_stats <- map_links_with_stats %>% 
    addPolylines(data = map_df, lng = ~longitude, lat = ~latitude, color = "green",
                 opacity = 1, label = ~linkid)
}

# Draw links with no stats in red
for (link in unique(links_no_stats$linkid)) {
  map_df <- links_no_stats %>% 
    filter(linkid == link)
  
  map_links_with_stats <- map_links_with_stats %>% 
    addPolylines(data = map_df, lng = ~longitude, lat = ~latitude, color = "red",
                 opacity = 1, label = ~linkid)
}

map_links_with_stats %>% 
  addCircles(data = dir1_stops, lat = ~stop_lat, lng = ~stop_lon, label = ~ stop_id, radius = 50, opacity = 1) %>%
  addCircles(data = dir2_stops, lat = ~stop_lat, lng = ~stop_lon, label = ~ stop_id, radius = 50, opacity = 1)
```

### Generate stats for missing links

This will be done by taking the average time/distance between the prior and post links per aggregate time, then multiplying by the distance of the missing links to get the estimated time on that link

```{r}
for(i in 1:nrow(missing_links_impute)) {
  link_length <- links %>% 
    filter(id == missing_links_impute$linkid[i]) %>% 
    pull(length)
  
  link_estimates <- link_stats %>% 
  filter(linkid %in% c(missing_links_impute$prior_linkid[i], missing_links_impute$post_linkid[i])) %>% 
  mutate(time_len = tt / length) %>% 
  group_by(logtime) %>% 
  summarise(
    avg_time_len = mean(time_len),
    direction = unique(direction)) %>% 
  ungroup() %>% 
  mutate(
    linkid = missing_links_impute$linkid[i],
    tt = round(avg_time_len * link_length),
    length = link_length) %>% 
  select(-avg_time_len)
  
  link_stats <- link_stats %>% 
  full_join(link_estimates, by = names(link_estimates))
}

# Remove estimator links not in segment
link_stats <- link_stats %>% 
  filter(linkid %in% c(links_dir1_enabled$linkid, links_dir2_enabled$linkid,
                       missing_links_impute$linkid))
```

Check

```{r}
links_with_stats <- links_segment %>% 
  filter(linkid %in% c(links_dir1_enabled$linkid, links_dir2_enabled$linkid,
                       missing_links_impute$linkid)) %>% 
  semi_join(link_stats, by = "linkid")

links_no_stats <- links_segment %>% 
  filter(linkid %in% missing_links_impute$linkid) %>%  
  anti_join(link_stats, by = "linkid")
  
map_links_with_stats <- leaflet() %>% 
  addTiles()

for (link in unique(links_with_stats$linkid)) {
  map_df <- links_with_stats %>% 
    filter(linkid == link)
  
  map_links_with_stats <- map_links_with_stats %>% 
    addPolylines(data = map_df, lng = ~longitude, lat = ~latitude, color = "green",
                 opacity = 1, label = ~linkid)
}

for (link in unique(links_no_stats$linkid)) {
  map_df <- links_no_stats %>% 
    filter(linkid == link)
  
  map_links_with_stats <- map_links_with_stats %>% 
    addPolylines(data = map_df, lng = ~longitude, lat = ~latitude, color = "red",
                 opacity = 1, label = ~linkid)
}

map_links_with_stats %>% 
  addCircles(data = dir1_stops, lat = ~stop_lat, lng = ~stop_lon, label = ~ stop_id, radius = 50, opacity = 1) %>%
  addCircles(data = dir2_stops, lat = ~stop_lat, lng = ~stop_lon, label = ~ stop_id, radius = 50, opacity = 1)
```

```{r}
# Calculate total links travel time per logtime and direction
links_times <- link_stats %>%
  group_by(logtime, direction) %>%
  summarise(links_time = sum(tt)) %>%
  ungroup() %>%
  mutate(
    day = date(logtime),
    time = as_hms(logtime),
    hour = hour(logtime),
    rush = case_when(
      between(time, as_hms("06:30:00"), as_hms("09:59:00")) ~ "Morning",
      between(time, as_hms("15:30:00"), as_hms("18:59:00")) ~ "Evening",
      TRUE ~ "Neither"),
    rush = factor(rush, levels = c("Morning", "Evening", "Neither"))) %>%
  relocate(day:rush, .after = logtime) %>% 
  select(-logtime)
```

# Analysis

## Compare bus travel times to links travel times to validate each other

```{r}
# Combine links times and bus trip times
travel_times <- trip_times_agg %>%
  inner_join(links_times) %>% 
  mutate(travel_diff = bus_time - links_time)

travel_times_piv <- travel_times %>% 
  pivot_longer(cols = bus_time:links_time, names_to = "type", values_to = "travel_time") %>% 
  mutate(type = recode(type, bus_time = "Bus", links_time = "Vehicle"))
```

```{r}
# Travel time distributions
travel_times_piv %>% 
  filter(rush != "Neither") %>% 
  ggplot(aes(travel_time/60, fill = type)) +
  geom_density(alpha = 0.4) +
  facet_grid(vars(direction), vars(rush)) +
  labs(x = "Travel Time (Minutes)", fill = "Type", title = "Bus Travel Time vs Vehicle Travel Time") +
  theme_bw()

# Direction 1 calender plot
travel_times_piv_dir1 <- travel_times_piv %>% 
  filter(direction == directions[1])

travel_times_piv_dir1 %>%
  ggplot(aes(time, travel_time, color = type)) +
  geom_line() +
  facet_wrap(vars(day), scales = "free_y") +
  scale_x_time(labels = label_time("%H:%M")) +
  labs(x = "Time",
       y = "Travel Time", 
       color = "Type", 
       title = "Northbound Bus Times vs Vehicle Times") +
  theme_bw() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.title = element_blank(),
    legend.justification = "bottom",
    legend.position = c(0.5, 0),
    legend.text = element_text(size = 10))

# Direction 2 calendar plot
travel_times_piv_dir2 <- travel_times_piv %>% 
  filter(direction == directions[2])

travel_times_piv_dir2 %>%
  ggplot(aes(time, travel_time, color = type)) +
  geom_line() +
  facet_wrap(vars(day), scales = "free_y") +
  scale_x_time(labels = label_time("%H:%M")) +
  labs(x = "Time",
       y = "Travel Time", 
       color = "Type", 
       title = "Southbound Bus Times vs Vehicle Times") +
  theme_bw() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.title = element_blank(),
    legend.justification = "bottom",
    legend.position = c(0.5, 0),
    legend.text = element_text(size = 10))

# Scatter plots
travel_times %>% 
  filter(rush != "Neither") %>% 
  ggplot(aes(bus_time/60, links_time/60)) +
  geom_point(alpha = 0.3) +
  scale_x_continuous(labels = label_comma()) +
  scale_y_continuous(labels = label_comma()) +
  labs(x = "Bus Travel Times (Minutes)",
       y = "Vehicles Travel Times (Minutes)",
       title = "Bus Travel Times vs Vehicles Travel Times per Direction per Rush Hour") +
  facet_grid(vars(direction), vars(rush), scales = "free") +
  theme_bw()
```

### Correlation between bus travel time and links travel time

```{r}
travel_times %>% 
  filter(rush != "Neither") %>% 
  group_by(rush, direction) %>% 
  summarise(Correlation = round(cor(bus_time, links_time), 2)) %>% 
  rename(Direction = direction,
         Rush = rush) %>% 
  kable(caption = "Correlation between travel times per direction per rush hour") %>% 
  kable_styling(full_width = FALSE)
```

## Average pattern of links time and bus trips time

```{r}
# Average total links time in each 30 minute bin
links_30min <- link_stats %>%
  group_by(logtime, direction) %>%
  summarise(links_time = sum(tt)) %>%
  ungroup() %>% 
  mutate(logtime = ceiling_date(logtime, "30 mins"),
         time = as_hms(logtime)) %>% 
  group_by(time, direction) %>% 
  summarise(links_time = mean(links_time)) %>% 
  ungroup() %>% 
  mutate(rush = case_when(
           between(time, as_hms("06:30:00"), as_hms("10:00:00")) ~ "morning",
           between(time, as_hms("15:30:00"), as_hms("19:00:00")) ~ "evening",
           TRUE ~ "neither"),
         rush = factor(rush, levels = c("morning", "evening", "neither"))) %>% 
  relocate(links_time, .after = everything())

# Average bus trip times in each 30 minute bin
trip_times_30min <- trip_times %>% 
  mutate(arrival_time = ceiling_date(arrival_time, "30 minutes"),
         time = as_hms(arrival_time)) %>% 
  group_by(time, direction) %>% 
  summarise(bus_time = round(mean(trip_time))) %>%
  ungroup() %>% 
  mutate(rush = case_when(
           between(time, as_hms("06:30:00"), as_hms("10:00:00")) ~ "morning",
           between(time, as_hms("15:30:00"), as_hms("19:00:00")) ~ "evening",
           TRUE ~ "neither"),
         rush = factor(rush, levels = c("morning", "evening", "neither"))) %>% 
  relocate(bus_time, .after = everything())

links_trips_30min <- links_30min %>% 
  inner_join(trip_times_30min) %>% 
  pivot_longer(cols = links_time:bus_time, names_to = "type", values_to = "value") %>% 
  mutate(type = recode(type, bus_time = "Bus", links_time = "Vehicle"))

ggplot(links_trips_30min, aes(time, value/60)) +
geom_line(alpha = 0.2, size = 1) +
geom_line(data = filter(links_trips_30min, rush == "morning"), color = "blue", size = 1) +
geom_line(data = filter(links_trips_30min, rush == "evening"), color = "blue", size = 1) +
facet_grid(rows = vars(type), cols = vars(direction)) +
annotate("rect", xmin = as_hms("06:30:00"), xmax = as_hms("10:00:00"), ymin = -Inf, ymax = Inf,
         fill = "red", alpha = .1) +
annotate("rect", xmin = as_hms("15:30:00"), xmax = as_hms("19:00:00"), ymin = -Inf, ymax = Inf,
         fill = "red", alpha = .1) +
scale_x_time(limits = as_hms(c("06:00:00", "23:59:59")),
             breaks = as_hms(c("06:00:00", "08:00:00", "10:00:00", "12:00:00",
                               "15:00:00", "17:00:00", "19:00:00")),
             labels = c("6", "8", "10", "12", "15", "17", "19")) +
scale_y_continuous(labels = label_comma()) +
labs(x = "Time (Hour)",
     y = "Travel Time (Minutes)",
     title = "Average Travel Time Pattern") +
theme_bw()
```

## Correlations

```{r}
standardiser <- function(basis = hour) {
  if (!deparse(substitute(basis)) %in% c("time", "hour", "rush")) {
    stop("basis must be either time, hour, or rush")
  }
  
  trip_times_std <- trip_times_agg %>%
    group_by({{basis}}, direction) %>% 
    mutate(
      period_buses_no = n(),
      bus_time_std = as.vector(scale(bus_time))) %>% 
    ungroup() %>% 
    mutate(bus_time_std = if_else(is.nan(bus_time_std), 0, bus_time_std)) %>%
    filter(abs(bus_time_std) < 3) %>%
    select(-bus_time)
    
  links_times_std <- links_times %>% 
    group_by({{basis}}, direction) %>% 
    mutate(links_time_std = as.vector(scale(links_time))) %>% 
    ungroup() %>% 
    filter(abs(links_time_std) < 3) %>%
    select(-links_time)
  
  # Combine
  travel_times_std <<- trip_times_std %>%
    inner_join(links_times_std) %>%
    mutate(std_travel_diff = bus_time_std - links_time_std)
  
  travel_times_std_piv <<- travel_times_std %>% 
    pivot_longer(cols = bus_time_std:links_time_std, names_to = "type", values_to = "std_travel_time") %>% 
    mutate(type = recode(type, bus_time_std = "Bus", links_time_std = "Vehicle"))
}

standardiser()
```

#### Visualise bus time and links time

```{r}
# Morning Northbound
travel_times_std_NB_morn <- travel_times_std %>% 
  filter(direction == "NB",
         rush == "Morning")

travel_times_std_piv_NB_morn <- travel_times_std_piv %>% 
  filter(direction == "NB",
         rush == "Morning")

travel_times_std_piv_NB_morn %>%
  ggplot(aes(time, std_travel_time, color = type)) +
  geom_line(size = 0.75) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(vars(day)) +
  scale_x_time(labels = label_time("%H:%M")) +
  labs(x = "Time",
       y = "Travel Time", 
       color = "Type", 
       title = "Standardized Northbound Morning Bus Times vs Vehicle Times") +
  theme_bw() +
  theme(
    legend.title = element_blank(),
    legend.justification = "bottom",
    legend.position = c(0.5, 0),
    legend.text = element_text(size = 10))

# Evening Southbound
travel_times_std_SB_eve <- travel_times_std %>% 
  filter(direction == "SB",
         rush == "Evening")

travel_times_std_piv_SB_eve <- travel_times_std_piv %>% 
  filter(direction == "SB",
         rush == "Evening")

travel_times_std_piv_SB_eve %>%
  ggplot(aes(time, std_travel_time, color = type)) +
  geom_line(size = 0.75) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(vars(day)) +
  scale_x_time(labels = label_time("%H:%M")) +
  labs(x = "Time",
       y = "Travel Time", 
       color = "Type", 
       title = "Standardized Southbound Evening Bus Times vs Vehicle Times") +
  theme_bw() +
  theme(
    legend.title = element_blank(),
    legend.justification = "bottom",
    legend.position = c(0.5, 0),
    legend.text = element_text(size = 10))
```

#### Travel time distributions

```{r}
travel_times_std_piv %>% 
  filter(rush != "Neither") %>% 
  ggplot(aes(std_travel_time, fill = type)) +
  geom_density(alpha = 0.4) +
  facet_grid(vars(direction), vars(rush)) +
  labs(x = "Standardized Travel Time",
       fill = "Type",
       title = "Standardized Bus Travel Time vs Vehicle Travel Time") +
  theme_bw()
```


#### Scatter plots

```{r}
travel_times_std %>% 
  filter(rush != "Neither") %>% 
  ggplot(aes(links_time_std, bus_time_std)) +
  geom_point(alpha = 0.3) +
  facet_grid(vars(direction), vars(rush)) +
  labs(x = "Standardized Bus Travel Times",
       y = "Standardized Vehicles Travel Times",
       title = "Bus Travel Times vs Vehicles Travel Times per Direction per Rush Hour") +
  theme_bw()
```

#### Correlation coefficients

```{r}
travel_times_std %>% 
  filter(rush != "Neither") %>% 
  group_by(rush, direction) %>% 
  summarise(Correlation = round(cor(bus_time_std, links_time_std), 2)) %>% 
  ungroup() %>% 
  rename(Direction = direction,
         Rush = rush) %>% 
  kable(caption = "Correlation between standardized travel times per direction per rush hour") %>% 
  kable_styling(full_width = FALSE)
```

